---
title: "Graphical data analysis 1"
output: html_document
---

```{r}
load("out.rData")
library(ggplot2)
```

## SubjectID

One subjectID for each patient

## Mineral balance

**Calcium** routine metabolic panel to assess kidney, bone, or nerve disease. Normal ranges are 2.2-2.5 mmol/L

**Phosphorus** Associated with kidney function, nutritional status, and a variety of chronic illnesses. Normal ranges are 1-1.5 mmol/L.

```{r}
summary(out$Calcium)

ggplot(na.omit(out), aes(na.omit(out$Calcium))) +
  geom_histogram(aes(y=..density..), binwidth=0.01, fill="lightblue") +
  geom_density() +
  geom_vline(xintercept=c(2.2,2.5)) +
  ylab("") +
  xlab(" Calcium level mmol/L")

ggplot(na.omit(out), aes(na.omit(out$Phosphorus))) +
  geom_histogram(aes(y=..density..), binwidth=0.01, fill="lightblue") +
  geom_density() +
  geom_vline(xintercept=c(1,1.5)) +
  ylab("") +
  xlab("Phosphorus level mmol/L")
```

## Blood count

**Red Blood Cells (RBC)** A count of the actual number of red blood cells per volume of blood. Both increases and decreases lead to abnormal conditions. Normal ranges are 4.2 - 6.9 $\times 10^9$/L.

[127] "RBC Distribution Width"                   
[128] "RBC Morphology"                           
[129] "RBC Morphology: Anisocytosis"             
[130] "RBC Morphology: Macrocytosis"             
[131] "RBC Morphology: Microcytosis"             
[132] "RBC Morphology: Spherocytes"              
[133] "Red Blood Cells (RBC)"     

```{r}
boxplot(out[,133], horizontal=TRUE, pch=5, xlab="Red Blood Cells (L)")
```

Substantially positive skewness is observed. So transformation methods will be applied and selected for further analysis.

```{r}
rbc <- out[,133]
trans1 = sqrt(rbc)
trans2 = log10(rbc)
trans3 = 1/rbc

par(mfrow=c(1,3))
qqnorm(trans1, main="Square root", pch=15)
qqnorm(trans2, main="Logarithmic", pch=15)
qqnorm(trans3, main="Reciprocal", pch=15)
```

## ALS Functional Rating Scale (ALSFRS)

The ALSFRS scale is a list of 10 assessments regarding motor function to assess symptom severity, with each measure ranging from 0 to 4, with 4 being the highest (normal function) and 0 being no function. The score for the individual questions are then summed together to generate a number, and that is the ALSFRS score.

In ALSFRS-R, the modified version of ALSFRS, assessment 10 (respiratory function) is further divided into three questions to better reflect the importance (weighting) of respiratory changes within the scale. 

[14] "ALSFRS_R _Total"                           
[15] "ALSFRS_Total" 

[111] "Q1_Speech"                                
[112] "Q10_Respiratory"                          
[113] "Q2_Salivation"                            
[114] "Q3_Swallowing"                            
[115] "Q4_Handwriting"                           
[116] "Q5_Cutting"                               
[117] "Q5a_Cutting _without _Gastrostomy"          
[118] "Q5b_Cutting _with _Gastrostomy"             
[119] "Q6_Dressing _and _Hygiene"                  
[120] "Q7_Turning _in _Bed"                        
[121] "Q8_Walking"                               
[122] "Q9_ Climbing_Stairs"   

```{r}
par(mfrow=c(1,1))
boxplot(out[,14], horizontal=TRUE, pch=5, xlab="Summed scores in ALSFRS assessments")
```

ALSFRS score is negatively skewed in a slight manner. So transformation methods will be applied and selected for further analysis.

```{r}
frs <- out[,14]
trans1 = frs^2
trans2 = frs^3
trans3 = frs^4

par(mfrow=c(1,3))
qqnorm(trans1, main="Squared", pch=15)
qqnorm(trans2, main="Cubed", pch=15)
qqnorm(trans3, main="To the fourth", pch=15)
```

## Lung function measures

**Forced vital capacity (FVC)** the volume of air that can forcibly be blown out after full inspiration, measured in liters. 

FVC Normal: the expected value for a non-ALS patient (control) matched by gender, age and height)

FVC percent: divide FVC liters by FVC normal

 [53] "fvc"                                      
 [54] "fvc_normal"                               
 [55] "fvc_percent"  

**Slow vital capacity** the maximum volume of air that can be exhaled slowly after slow maximum inhalation, measured in liters.

[142] "svc"                                      
[143] "svc_normal"                               
[144] "svc_percent" 

```{r}
summary(out$fvc_percent)
summary(out$svc_percent)
```


## Treatment group

Some patients received placebo treatments, while others received experimental treatments (medication).

```{r}
out$treatment_group[is.na(out$treatment_group)] <- "Unknown"
lv <- levels(factor(out$treatment_group))

ggplot(out, aes(out$fvc_percent, fill=out$treatment_group)) +
  geom_density(alpha=.6)
```

